---
title: "Global Terrorism Visualization"
author: "Bryce Curtsinger"
date: "December 11, 2018"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

In this exercise, we will plot geospatial data in R using ggplot2 and ggmap. The data is a list of terrorist attacks worldwide since the 1970s. For obvious reasons, this post will be less personal than my other two blog posts, but the data provides a large number of observations and several dimensions we can consider to get a wide variety of plots. This post will highlight several features of ggplot2 that are useful in any plotting situation, as well as mapping-specific features. Familiarity with basic ggplot functions and structure is recommended to gain the most knowledge from this post.

Below I provide the data source & a brief description of packages used in this analysis.

Data source: https://www.kaggle.com/START-UMD/gtd

pacman - easy installation / loading of packages

tidyverse - data manipulation toolkit

ggplot2 - plotting toolkit

ggmap - mapping toolkit that integrates with ggplot2

```{r Intro & Data Prep, echo = F, message=F}
if(!require(pacman)) install.packages('pacman')
p_load(tidyverse)
p_load(ggplot2)
p_load(ggmap)
gtd <- read_csv('gtd.csv')

#Subset Data
gtd <- gtd %>% select(eventid, iyear,imonth,iday,country_txt,region_txt,provstate,city,latitude,longitude,attacktype1_txt, targtype1_txt, gname, weaptype1_txt, nkill, nkillus) %>% filter(!is.na(latitude), !is.na(longitude), longitude > -500, !is.na(nkill))
#Summarize
summary(gtd)
```

## Visualizing Terrorism

To begin, let's consider attacks split by decade. We will plot each incident separately, with color and size decided by the deadliness (measured as number of fatalities). we can initialize a world map using the "borders" function in ggmap, specifying 'world' as the database. Border and fill colors can be customized as well, but let's stick with a simple gray background with black borders. We can even save this base plot to an object, so we don't have to specify it each time we want to create a new map.

In each plot, we can use tidyverse grammar to filter our data by decade in the ggplot functions. We will add points for each incident with geom_point, then add a color gradient to our points with scale_color_gradient, ranging from a light blue to red for low to high deadliness.

```{r Visualize Terror Incidents}
#Set up world map:
mapWorld <- borders("world", colour="black", fill="gray70")
wm <- ggplot(gtd, aes(x = longitude, y = latitude)) + mapWorld

#Plot by Decade

#70s
wm + geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(iyear >= 1970 & iyear <= 1980 & nkill <= 200)) + 
  geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(iyear >= 1970 & iyear <= 1980 & nkill > 200)) + 
  scale_color_gradient(low = "steelblue3", high = 'red', na.value = "grey50", guide = "colourbar" )
#80s
wm + geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(iyear >= 1980 & iyear <= 1990 & nkill <= 200)) + 
  geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(iyear >= 1980 & iyear <= 1990 & nkill > 200)) + 
  scale_color_gradient(low = "steelblue3", high = 'red', na.value = "grey50", guide = "colourbar" )
#90s
wm + geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(iyear >= 1990 & iyear <= 2000 & nkill <= 200)) + 
  geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(iyear >= 1990 & iyear <= 2000 & nkill > 200)) + 
  scale_color_gradient(low = "steelblue3", high = 'red', na.value = "grey50", guide = "colourbar" )
#2000s
wm + geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(iyear >= 2000 & iyear <= 2010 & nkill <= 200)) + 
  geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(iyear >= 2000 & iyear <= 2010 & nkill > 200)) + 
  scale_color_gradient(low = "steelblue3", high = 'red', na.value = "grey50", guide = "colourbar" )
#2010s
wm + geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(iyear >= 2010 & iyear <= 2020 & nkill <= 200)) + 
  geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(iyear >= 2010 & iyear <= 2020 & nkill > 200)) + 
  scale_color_gradient(low = "steelblue3", high = 'red', na.value = "grey50", guide = "colourbar" )

```

## Facets
An alternative approach to 5 separate graphs is to use facets included in ggplot. Below we will reproduce these plots using a facet. First, a variable indicating the decade should be created, then used as the argument in the facet_wrap function.

```{r facet}
gtd <- gtd %>% mutate(decade = ifelse(iyear < 1980, '1970s', ifelse(iyear < 1990, '1980s', ifelse(iyear < 2000, '1990s', ifelse(iyear < 2010, '2000s', '2010s')))))

wm + geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(nkill <= 200)) + 
  geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(nkill > 200)) + 
  scale_color_gradient(low = "steelblue3", high = 'red', na.value = "grey50", guide = "colourbar" ) + 
  facet_wrap(~decade)

```


We see the same plots produced above in a nicely arranged grid. Additionally, this approach will standardized the size and color scales on all plots to make comparing across decades much simpler. We can see the spread of incidents and the magnitude of deadliness change over time.

Let's consider another use of the facet_wrap function, this time considering attack type instead of decade.

```{r facet II}
wm + geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(nkill <= 200)) + 
  geom_point(aes(x = longitude, y = latitude, color = nkill, size = nkill), data = gtd %>% filter(nkill > 200)) + 
  scale_color_gradient(low = "steelblue3", high = 'red', na.value = "grey50", guide = "colourbar" ) + 
  facet_wrap(~attacktype1_txt)
```

Facets in ggplot make comparing across categorical measures quick and easy. As you can see with the plot above, the plots can become hard to read when wrapping across categorical variables with a lot of levels.

## Specialized Visualizations

In the plots above, it can be difficult to identify how many incidents occur in an area due to overlapping points and the sheer number of datapoints. One way to effectively visualize the density of points on a plot is with a heatmap. Below, we will use the "stat_density2d" function in ggplot to generate a heatmap. Note the argument to the 'fill' command "..level.." will automatically scale the color based on the density at that location, and 'alpha' = 0.5 decides how see-through our heatmap is.

```{r Heatmap}
wm  + stat_density2d(data=gtd, aes(x=longitude, y=latitude, fill=..level.., alpha = 0.5), geom="polygon")
```

We can even combine our heatplot maps with the facet_wrap function we used above. This provides a great visualization of how terrorism "hotspots" changed over time.

```{r Facet Heatmap}
wm  + stat_density2d(data=gtd, aes(x=longitude, y=latitude, fill=..level.., alpha = 0.5), geom="polygon") + facet_wrap(~decade)
```

## Conclusions & Takeaways

Working with geospatial data can be overwhelming, but knowing the right graphing tools can make visualizing this data a breeze. To review, we learned how to adapt basic ggplot2 grammar to include geospatial data with latitudes and longitudes, used facets to easily compare graphs across other measures, and created specialized visualizations in the form of heat maps to show the density of points on a map. Many other tools are available to improve the quality of maps and plot locations at higher levels of granularity. Some functions even allow you to integrate with Google Maps (assuming you have an API key) to pull satellite graphs or gps-quality maps into R. However, understanding the basics of mapping in R is necessary to move to this level of detail.